03_distances

### 2020-05-05 03_distances.Rmd
Recompute euclidean distances
This computes Euclidean distances for block groups and tract centroids to park boundaries. The file is now too large to keep in Git.

---
title: "Distances"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(reticulate)
library(tidyverse)
library(sf)
library(leaflet)
use_condaenv("ox", conda = "~/opt/anaconda3/condabin/conda", required = TRUE)
```

We want to calculate network-based distances between blocks in New York City
and the edges of parks.  In this document, we do the following:

  - Obtain a highway network for New York City from OpenStreetMap
  - Determine coordinates for block group centroids and park edge nodes
  - Compute the shortest path between the centroid and the nearest edge node.

To do this, we are going to use the `osmnx` library for  python.

```{python setup}
import osmnx as ox
import networkx as nx

import numpy as np
import pandas as pd
import time
```

## OpenStreetMap

We use the `osmnx` library to get a highway network graph for New York City from 
OpenStreetMap. We get the walk network because most people in New York City will
not be driving to get to their park.

```{python graph}
graph = ox.graph_from_place('New York, New York, USA', network_type='walk')
graph_proj = ox.project_graph(graph)
```

## Block Groups

We get a list of population-weighted block group centroids from Census.

```{python blockgroups}
# get the coordinates from census
url = "https://www2.census.gov/geo/docs/reference/cenpop2010/blkgrp/CenPop2010_Mean_BG36.txt"
blockgroups = pd.read_csv(url)

# pad the county and tract strings, and get the proper data types
blockgroups.STATEFP = blockgroups.STATEFP.apply(str)
blockgroups.BLKGRPCE = blockgroups.BLKGRPCE.apply(str)
blockgroups.COUNTYFP = blockgroups.COUNTYFP.apply('{0:0>3}'.format)
blockgroups.TRACTCE = blockgroups.TRACTCE.apply('{0:0>6}'.format)
blockgroups['GEOID'] = blockgroups[['STATEFP', 'COUNTYFP', 'TRACTCE', 'BLKGRPCE']].apply(lambda x: ''.join(x), axis = 1)

blockgroups['LATITUDE'] = blockgroups['LATITUDE'].astype(float)
blockgroups['LONGITUDE'] = blockgroups['LONGITUDE'].astype(float)

# filter to nyc
nyc_counties = ["081", "047", "061", "005", "085"]
blockgroups = blockgroups[blockgroups.COUNTYFP.isin(nyc_counties)]
blockgroups.set_index('GEOID', inplace=True)
```

## Open Spaces

We want to calculate distance from the block group centroid to the nearest
edge of each open space. The open spaces data created on the 
[previous page](02_parks_data.html) is a detailed simple features layer with a
polygon for each open space. The `osmnx` graph needs to have point coordinates 
to calculate network-based distances, however. To make this work, we

 1. simplify the polygon geometry to reduce the number of edge vertices. We do this
 with a tolerance of 100 feet.
 2. convert the polygon geometry into a boundary linestring
 3. sample points on the linestring with a density of 1 point every 500 feet. Parks
 with a permeter of less than 500 feet will have a single point.

```{r simplify_spaces}
# get open spaces file
open_spaces <- read_rds("data/open_spaces.rds") %>%
  select(id)  %>%
  # some polygons are way too detailed, so we want to simplify to 100 foot resolution
  st_simplify(dTolerance = 100, preserveTopology = TRUE)

# some parks have multiple polygons. We need these to be simpler individual features 
poly_from_multi <- open_spaces %>%
  filter(st_geometry_type(geometry) == "MULTIPOLYGON") %>%
  st_cast("POLYGON")

linestrings <- open_spaces %>%
  # join the original polygons back with the simplified multipolygons
  filter(st_geometry_type(geometry) != "MULTIPOLYGON") %>%
  rbind(poly_from_multi) %>%
  # cast the polygons to a linestring of the park perimeter
  st_cast("LINESTRING", group_or_split = TRUE)

point_samples <- linestrings %>%
  # sample points along the line, one point per 1000 feet.
  st_line_sample(density = 1/1000)

# append open space id
open_space_points <- tibble(id = linestrings$id, geometry = point_samples) %>%
  st_as_sf() %>%
  st_cast(to = "POINT")
```

The results of this data munging are given below. We have some issues where parks 
made of multiple polygons end up with internal points. Trying to get rid of these
by using complex hulls created other problems on linear parks. So we'll just go
with this.

```{r simplified-point-map}
leaflet() %>%
  addProviderTiles(providers$OpenStreetMap) %>%
  addCircleMarkers(data = open_space_points %>% st_transform(4326), radius = 0.001,
                   color = "red") %>%
  addPolygons(data = open_spaces %>% st_transform(4326))
```

Get the data into a form that we can pass it into python.

```{r park_points}
park_points <- open_space_points %>%
  st_transform(4326) %>%
  mutate(
    LONGITUDE = st_coordinates(.)[, 1],
    LATITUDE = st_coordinates(.)[,2]
  ) %>%
  st_set_geometry(NULL) %>%
  as_tibble()
```

## Distances

First, we bring the park points we sampled into a pandas data frame with
the park id as the index.
```{python park_points_py}
park_points = pd.DataFrame(r.park_points)
# TODO: run for all park points
park_ids = park_points.sample(2).id

park_points.set_index('id', inplace=True)
park_points = park_points.loc[park_ids]

```

We already created the block group centroids data frame earlier. Let's get the
nearest network nodes for each of the block groups and each park point.

```{python park_bg_nodes}
# TODO: run for all block groups
blockgroups = blockgroups.sample(5)

def findnode(df):
  df['node'] = df.apply(lambda row: 
    ox.get_nearest_node(graph, (row.LATITUDE, row.LONGITUDE)), axis=1)
  return df

a = time.process_time()
findnode(blockgroups)
findnode(park_points)
b = time.process_time()
b - a
```

We want to create a data structure that we can quickly populate with the 
shortest path between a block group and the nearest park point. This is 
probably a two-level dictionary.

```{python blockgroup_dict}
# Create a dictionary for the block groups
blockgroup_dict = {}
for i in blockgroups.index.unique():
  blockgroup_dict[i] = {} # create a dictionary for the parks
  for j in park_points.index.unique():
    blockgroup_dict[i][j] = float("inf")

```

Now we want to loop through each block group and all of the park points to figure
out what the shortest path is.

```{python nearest}
# loop through block groups
a = time.process_time()
for bg in blockgroups.itertuples():

  # loop through parks
  for park in park_ids:
    these_points = park_points.loc[[park]]

    # loop through points
    for point in these_points.itertuples():
      # get length between park and this edge point
      try:
        length = nx.shortest_path_length(graph_proj, source = bg.node, target=point.node,
          weight = 'length')
      except:
        length = float("inf")

      # check if this is the shortest one we've found; if so, update.
      if length < blockgroup_dict[bg.Index][park]:
        blockgroup_dict[bg.Index][park] = length

b = time.process_time()
b - a 
```

```{python write}
park_points.to_csv("data/park_points.csv")
blockgroups.to_csv("data/blockgroups.csv")
```

>>>>

---
title: "03_distances"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(sf)
counties <- c("081", "047", "061", "005", "085")
```

A key aspect of accessibility to parks is the distance between parks and
neighborhoods. There are two different ways to calculate distance:

   - *Euclidean* distances use the straight-line distance between two points.
   - *Network* distances use the path along a highway or sidewalk network.

Both measurements have limitiations. Euclidean distances are particularly
problematic when there are rivers, interstates, or other impassible features that
substantially lengthen the distance people travel to get between points. On the
other hand, there is not lots of data suggesting that people perceive network 
travel times very well when they are choosing destinations. We originally
estimated the models for this paper with Euclidean distances, but
our reviewers suggested we examine network-based distances as well.

Another issue the reviewers identified is the spatial resolution of the analysis.
We had used tracts, because this is where the socioeconomic and health data are 
available. But block groups are more geographically precise, and it is possible
for us to compute the accessibility at that level, if we assume that the 
socioeconomics and health information is constant across the tract (this is not
a good assumption, but we can entertain it for the purposes of the response to
reviewers).

We will use the same parks dataset we assembled [previously](02_parks_data.html).

```{r open_spaces}
open_spaces <- read_rds("data/open_spaces.rds")
```

## Euclidean Distances

The geometric centroid of the tract / block group is not a good approximation of
the center point of the tract from an accessibility standpoint. Thus, we 
get the population-weighted centroid for tracts and block groups from Census.

```{r tract_centroids}
tracts_url <- "https://www2.census.gov/geo/docs/reference/cenpop2010/tract/CenPop2010_Mean_TR36.txt"

tract_centroids <- read_csv(tracts_url) %>%
  filter(COUNTYFP %in% counties) %>%
  transmute(geoid = str_c(STATEFP, COUNTYFP, TRACTCE), LATITUDE, LONGITUDE) %>%
  # convert to sf and project into the feet-based projection used in the
  # parks dataset.
  st_as_sf(crs = 4326, coords = c("LONGITUDE", "LATITUDE")) %>%
  st_transform(3628) 
```

```{r bg_centroids}
bg_url <- "https://www2.census.gov/geo/docs/reference/cenpop2010/blkgrp/CenPop2010_Mean_BG36.txt"

bg_centroids <- read_csv(bg_url) %>%
  filter(COUNTYFP %in% counties) %>%
  transmute(geoid = str_c(STATEFP, COUNTYFP, TRACTCE, BLKGRPCE), LATITUDE, LONGITUDE) %>%
  # convert to sf and project into the feet-based projection used in the
  # parks dataset.
  st_as_sf(crs = 4326, coords = c("LONGITUDE", "LATITUDE")) %>%
  st_transform(3628) 
```

The `st_distance()` function from the `sf` library is able to compute the 
Euclidean distance between two sets of shapes, and gets the distance between 
points and edges of polygons directly. We calculate the distance in miles, 
and constrain that the minimum distance is 0.1 miles (no one lives in a park).

```{r distance_table}
distance_table <- function(points, poly){
  dists <- st_distance(points, poly, by_element = FALSE) %>%
    units::set_units(miles) %>%
    units::drop_units()

  dists <- pmax(dists, 0.1)

  rownames(dists) <- points$geoid
  colnames(dists) <- poly$id

  as_tibble(dists, rownames = "geoid") %>%
    pivot_longer(cols = -geoid, names_to = "park_id", values_to = "distance")
}

dist_tr_euc <- distance_table(tract_centroids, open_spaces)
dist_bg_euc <- distance_table(bg_centroids,    open_spaces)
```

```{r distances}
list("tracts" = dist_tr_euc, "blockgroups" = dist_bg_euc) %>%
  write_rds("data/distances.rds")
```

## Network Distances

For this
